spinifex: manual control of dynamic linear projections of high-dimensional data

### 2019-04-09 paper/spinifex-paper.tex
remainder of changes to DIS. Grammerly and knit. add a wrap up summary.

\abstract{%
The class of dynamic linear projections that are collectively known as `tours' provide a unique dynamic visualization of numeric multivariate data. Tours are particularly useful for understanding the structure held within multivariate data, and in association with techniques for dimension reduction, supervised, and unsupervised classification. The \emph{R} package \emph{tourr} offers a variety of path generators and geometric displays for conducting tours. This paper discusses an extension package, \emph{spinifex}, that adds support for the path generation of manual tours and extends the display of tours to use with the contemporary animation packages, \emph{plotly} and \emph{gganimate}. Manual tours are used to explore the sensitivity of structure as the contributions of a manipulation variable are changed. This particularly useful after identifying a feature of interest.
A recent paper \{Wang et al. (2018)\} visualizes the sensitivey of the hadronic experiments to nucleon structure. Sensitivity was characterized in non-linear 3D embeddings of the first 10 principal components. This research applies manual tours to this data showing that manual tours resolves more structrual information that is orthogonal to the original viewing plane.
}

@@

A tour is a multivariate data analysis technique in which a sequence of linear (orthogonal) projections are viewed as an animation while the orientation of the projection basis is rotated across time. Each frame of the sequence corresponds to a small change in the projection for a smooth transition that perseveres continuity.

While there are numerous methods that generate tour paths, this research focuses on the manual tour. The manual tour was described in Cook and Buja (1997) and allows a user to control the projection coefficients of a select variable has in a 2D projection. The manipulation of these coefficients allows the analyst to explore how sensitive the projections structure is to these changes. This makes manual tours particularly useful once a feature of interest has been identified, for example, with the use of a guided tour (Cook et al. 1995). The path of a guided tour is selected via projection pursuit, the optimization of an index function on the projection via a hill climbing algorithm. This allows guided tours to identify interesting projection features rapidly given the relatively large parameter-space. Once the given projection has been provided, it is time to define the path of rotation.

Ideally, the path would be intuitively user-generated from physical movement, be it through mouse or motion capture. Unfortunately, this type of dynamic control has proven difficult to capture for in R. Because of this, manual tours were not implemented in \emph{tourr}. This research allows for the consumption, but not the generation, of such dynamic input. After the capture of an oblique user motion, the rotation needs to be applied to step 3 (rotation sequence) of the algorithm discussed below. In the section below we stick with a radial rotation where, \(\theta\), the angle of in-projection-plane rotation is held constant.

Spinifex utilizes two new animation packages, \emph{plotly} (Sievert 2018) and \emph{gganimate} (Pedersen and Robinson 2019), to display tours, manual or other saved tours. From a given projection, the user can choose which variable to control, and the animation sequence is generated to remove the variable from the projection, and then extend its contribution to be the sole variable in one direction. This allows the viewer to assess the change in structure induced in the projection by the variable's contribution.

@@

\begin{itemize}
  \item $\textbf{X}$, the data, an $n \times p$ numeric matrix to be embedded in two dimensions.
  \item $\textbf{B} = (B_1,~ B_2)$, any orthonormal projection basis set, $p \times 2$ matrix, describing the projection from $p$ to two dimensions.
  \item $\textbf{e}$, a zero column vector of length $p$ with the $k-$th element set to one, where $k$ is the number of the variable to manipulate.
  \item $\theta$, the angle of in-projection-plane rotation, for example, on the reference axes. 
  \item $\phi$, the angle of out-of-projection-plane rotation, coming into the manipulation space.
\end{itemize}

The algorithm primarily operates on the projection basis and utilizes the data only when making a display. The projection space can be viewed at any point in the process by pre-multiplying the data and plotting the first 2 variables.

@@

The flea data, originally from Lubischew (1962), available in the R package \emph{tourr} (Wickham et al. 2011) is used to illustrate the algorithm. The data contains 74 observations across 6 variables, physical measurements of the flea beetles. Each observation belonging to one of three species.

The data is defined. A basis set (ideally that views an interesting feature) should be provided to explore the sensitivity of the variables to the structure. To identify a projection containing an interesting feature, apply
a guided tour(Cook, Swayne, and Buja 2007) on the flea data. In a guided tour the projection sequence is selected by optimizing an index via hill-climbing. In this case, the holes index is selected. The holes index is maximized by when the projected observations are furthest from the center. Figure @ref(fig:step0) shows a locally optimized projection for this data. The left plot displays the reference axes of the projection basis, a visual indication of the magnitude and direction each variable contributed to the projections. The right plot shows the projection of the data through the basis set described by the reference axes (left). Data points are colored and given point characters according to the species of the flea (the guided tour was unsupervised with this information).

@@

{\centering \includegraphics[width=0.95\linewidth]{spinifex_paper_files/figure-latex/step0-1}

@@

(ref:step0-cap) Basis reference axes (left) and projected data (right) of standardized flea data. Data points color and shape are mapped to beetle species. Basis identified by a holes-index guided tour. The variables \texttt{aede2} and \texttt{tars1} contribute mostly orthogonal to the other variables. We'll select \texttt{aede2} as our manipulation variable to see how the structure of the projection changes as we rotate \texttt{aede2} into and out of the projection.

Call \texttt{view\_basis()} on a basis to produce a \emph{ggplot2} graphic similar to @ref(fig:step0). Projection space is always available for display via the matrix multiplication \(\textbf{X}_{[n,~p]} ~*~ \textbf{B}_{[p,~d]} ~=~ \textbf{P}_{[n,~d]}\).

@@

In figure @ref(fig:step0), above, the contributions of the variables \texttt{tars1} and \texttt{aede2} are mostly orthogonal to the contributions of the other four variables. These two variables explain the variation of the data between the purple and green species. We select \texttt{aede2} as the manip var, the variable to be manipulated as it typically has a larger contribution after the optimizing the holes index. The question that will be explored in the explanation of the algorithm is how important the variable \texttt{aede2} is to the separation of the clusters.

@@

Initialize a zero vector \(e\) of \(p\) elements. Because \texttt{aede2} is the fifth variable in the data, set the \(k=5\)-th element to one giving the manip var a full contribution in this dimension. Use the Gram-Schmidt process to orthonormalize the zero vector onto the basis yielding the 3D manipulation space, \textbf{M}.

\begin{align*}
  \textbf{e} &\leftarrow Orthonormalize_{GS}(\textbf{e}) w.r.t. Basis \\
  &= \textbf{e} - \langle \textbf{e},\textbf{B}_1 \rangle \textbf{B}_1 - \langle \textbf{e}, \textbf{B}_2 \rangle \textbf{B}_2 \\
  \\
  \textbf{M}_{[p,~3]} &= (\textbf{B}_1,\textbf{B}_2,\textbf{e})
\end{align*}

Adding this extra dimension to our basis plane allows for the coefficients of the specified variable to be changed. For example, the ability to lift a piece of paper, rather than being constrained to the motion on a table top. Orthonormalizing rescales the new depth vector while the projection down to 2D is the original basis, that is the first \(d\) vectors remain constant. Imagine the reference axes (and projection plane) laying flat on a table, while a new dimension exists with axes projecting back onto the reference axes. An illustration of such can be seen below in figure @ref(fig:step2). The manip var is highlighted, while the depths of the other variables are not depicted.

(ref:step2-cap) Manipulation space for controlling the contribution of \texttt{aede2} of standardized flea data. Basis selected by a holes-index guided tour. The Projection plane is shown in blue. The manipulation axis, in red, allows the coefficients of the manip var to be changed.

@@

{\centering \includegraphics[width=1\linewidth]{spinifex_paper_files/figure-latex/step2-1}

@@

The representation in @ref(fig:step2) can be duplicated by calling the function \texttt{view\_manip\_space()}.

@@

Imagine holding the red axis it is fixed to the origin. As it is manipulated the projection back onto the projection plane correspondingly moves. This is what happens in a manual tour. For a radial tour, fix \(\theta\), the angle within the blue plane, and vary the sequence of \(\phi\), the angle coming out of the projection plane. Conceptually, live manipulation on a 2D plane allows the user to dynamically control these angles, effectively changing the coefficients of the manip var, which then performs a constrained rotation on the remaining variables.

For the demonstration of the radial tour, we define a sequence for \(\phi\) that brings the initial contribution of the manip var to be maximized and then zeroed before returning to the initial position.

@@

\begin{description}
  \item[$c_\theta$] be the cosine of $\theta$
  \item[$c_\phi$]   be the cosine of $\phi$
  \item[$s_\theta$] be the sine of   $\theta$
  \item[$s_\phi$]   be the sine of   $\phi$
\end{description}

then
\begin{align*}
  \textbf{RM}_{[p,~3,~i]}
  &= \textbf{M}_{[p,~3]} ~*~ \textbf{R}_{[3,~3]} \\
  &= \begin{bmatrix}
    M_{1,~1} & M_{1,~2} & M_{1,~3} \\
    M_{2,~1} & M_{2,~2} & M_{2,~3} \\
    \vdots   & \vdots   \\
    M_{p,~1} & M_{p,~2} & M_{p,~3}
  \end{bmatrix}_{[p,~3]}
    ~*~
  \begin{bmatrix}
    c_\theta^2 c_\phi s_\theta^2 &
    -c_\theta s_\theta (1 - c_\phi) &
    -c_\theta s_\phi \\
    -c_\theta s_\theta (1 - c_\phi) &
    s_\theta^2 c_\phi + c_\theta^2 &
    -s_\theta s_\phi \\
    c_\theta s_\phi &
    s_\theta s_\phi &
    c_\phi
  \end{bmatrix}_{[3,~3]}
\end{align*}

A note on application: compile the sequence of \(\phi_i\) and create an array/long table for each rotated manipulation space. \(\phi\) is the angle relative to the initial value of \(\phi\), we find the transformation \(\phi_i\) - \(\phi_1\) useful to think about \(\phi\) relative to the basis plane. Additionally, the value of \(\phi\) may be offset by a factor of pi. If the manip variable doesn't move as expected these are the first places to check.

\begin{Schunk}
\begin{Sinput}
for (phi in seq(seq_start, seq_end, phi_inc_sign)) {
  slide <- slide + 1
  tour[,, slide] <- rotate_manip_space(manip_space, theta, phi)[, 1:2]
}
\end{Sinput}
\end{Schunk}

Figure @ref(fig:step3) illustrates a sequence with 15 projected bases and highlight the manip variable on top while showing the corresponding projected data points on the bottom. Take note of how the changes in the manip var change the distance between the purple and green cluster of points, \texttt{aede2} is crucial in distinguishing between these groups. Tours are typically viewed as an animation such a dynamic version of this tour can be viewed online at \url{https://nspyrison.netlify.com/thesis/flea_manualtour_mvar5/}. The page may take a moment to load. The format of this figure and linking to an HTML animation will be used again in the Application, section \ref{sec:application}.

(ref:step3-cap) Radial manual tour changing the contributions from \texttt{aede2} of standardized flea data. The contributions increase from its initial contribution to a full contribution to the projection before decreasing to zero and then returning to its initial value. The change in the projected data shows that \texttt{aede2} is important for distinguishing between the purple and green clusters. An animated version can be viewed at \url{https://nspyrison.netlify.com/thesis/flea_manualtour_mvar5/}.

@@

Animations can be produced using the function \texttt{play\_manual\_tour()}. This function defaults to an HTML5 widget produced from \emph{plotly}.

@@

In light of performance, the above operations are performed on the bases without the use of the larger datasets. After the bases are brought into the projection-space, however, it is helpful to observe them with data in the same space. Pre-multiply the data by basis frame bringing the data into the projection space.

@@

The \emph{tourr} package utilizes R's base graphics for the display of tours. \emph{spinifex} allows tours to be used in rendered in \emph{plotly} Sievert (2018) as an HTML5 object or \emph{gganimate} Pedersen and Robinson (2019) as .gif or .mp4 objects. Both of which build off \emph{ggplot2} objects in internal functions. Sharing of animations is not trivial especially in print and static formats such as .pdf. Even with the use of computers and dynamic file formats capturing the correct resolution, aspect, and display is challenging and many formats quickly bloat file sizes. Keep in mind hosting options and exporting functions from \emph{plotly}, \emph{gganimate} and \emph{tourr}.

@@

Storing each data point for every frame of the animation is very inefficient. Just as operations are performed on the bases, so too should tour paths be stored as bases. Consider a radial manual tour, we can store the salient features in 3 bases, where \(\phi\) is at its starting, minimum, and maximum values. The frames in between can be interpolated by supplying angular speed. With the use of the \texttt{tourr::save\_history()} function, the target bases can be saved. From there geodesic interpolation can be used to populate the intermittent frames. This type of interpolation should not be used on manual tours, which have already been initialized into a 3D manipulation space where direct linear interpolation is appropriate.

@@

In a recent paper, Wang et al. (2018), the authors aggregate and visualize the sensitivity of hadronic experiments to ncleon structure. The authors introduce a new tool, PDFSense, to aid in the visualization of Parton distribution functions (PDF). The parameter-space of these experiments lies in 56 dimensions, \(\delta \in \mathbb{R}^{56}\), and are visualized as 3D subspaces of the 10 first principal components in linear (PCA) and non-linear (t-SNE) embeddings.

Using the same data, another study, Cook, Laa, and Valencia (2018), applies grand tours to the same subspaces. Grand tours are able to better resolve the distribution shape of clusters, intra-cluter detail, better outlier detection, and exonerate a claim persened from TFEP (TensorFlow embedded projections). Table 1 of Cook et al.~summarizes the key findings of observations made with PDFSense \& TFEP and those from grand tours.

Without getting too domain-specific the data has three primary groupings; DIS, VBP, and jet Each group is a particular class of experiments and each with many experimental datasets which inturn have many observation. Inconsideration of data density and business of the data We conduct manual tours on a subsets of the DIS and jet clusters. This explores the sensitivity of the structue to each of the variables in turn, and we present the subjectively best and worst manaul tour identifying structure in the respective data sets.

@@

The jet cluster is of interest as it contains the largest data sets and is found to be important in Wang et al. (2018) The jet cluster resides in a smaller dimensionality than the full set of experiments with 4 principal components explaining 95\% of the variation in the jet cluster (Cook, Laa, and Valencia 2018). The data is subset down to ATLAS7old and ATLAS7new to focus in on two groups with a reasonable number of observations that occupy different parts of the subspace. Below, we perform radial manual tours all four principal components within this scope. Visualizing PC3 and PC4 in figure @ref(fig:JetClusterGood) (more sturcturally insightful) and figure @ref(fig:JetClusterBad) (less sturcturally insightful) respectively, and list links to dynamic animation of all variables.

(ref:JetClusterGood-cap) Jet cluster, a radial manual tour of PC3. Colored by experiment type: `ATLAS7new' in green and `ATLAS7old' in orange. When PC3 fully contributes to the projection ATLAS7new (green) occupies unique space and several outliers are identifiable. Zeroing the contribution from PC3 to the projection hides the outliers and indeed all observations with ATLAS7new are contained within ATLAS7old (orange). A dynamic version can be viewed at \url{https://nspyrison.netlify.com/thesis/jetcluster_manualtour_pc3/}.

@@

(ref:JetClusterBad-cap) Jet cluster, a radial manual tour of PC4. Colored by experiment type: ATLAS7new in green and ATLAS7old in orange. This manual tour contains less interesting information ATLAS7new (green) has points that are right and left of ATLAS7old, while most points occupy the same projection space, regardless of the contribution of PC4. A dynamic version can be viewed at \url{https://nspyrison.netlify.com/thesis/jetcluster_manualtour_pc3/}.

@@

\hypertarget{todo-edit-application-grammarly-and-word-checks.}{%
\subsection{TODO: edit application, Grammarly and word checks.}\label{todo-edit-application-grammarly-and-word-checks.}}

Manipulating PC3, where varying the angle of rotation brings interesting features into and out of the center mass of the data, is more interesting than the manipulation of PC4, where the features are mostly independent of the contribution of PC4.

Jet cluster manual tours manipulating each of the principal components can be viewed from the links: \href{https://nspyrison.netlify.com/thesis/jetcluster_manualtour_pc1/}{PC1}, \href{https://nspyrison.netlify.com/thesis/jetcluster_manualtour_pc2/}{PC2}, \href{https://nspyrison.netlify.com/thesis/jetcluster_manualtour_pc3/}{PC3}, and \href{https://nspyrison.netlify.com/thesis/jetcluster_manualtour_pc4/}{PC4}.

@@

We perform a manual tour on this data, manipulating PC6 as depicted in figure @ref(fig:DISclusterGood). Looking at several frames we see that DIS HERA data lies mostly on a plane. When PC6 has full contributions, we see the dimuon SIDIS in purple is almost orthogonal to the DIS HERA (green). Yet the contribution of PC6 has zeroed the dimuon SIDIS data occupy the same space as the DIS HERA data. A dynamic version of this manual tour can be found at:
\url{https://nspyrison.netlify.com/thesis/discluster_manualtour_pc6/}.
The page may some time to load, as the animation is several megabytes.

(ref:DISclusterGood-cap) DIS cluster, a radial manual tour of PC6. colored by experiment type: `DIS HERA1+2' in green, `dimuon SIDIS' in purple, and `charm SIDIS' in orange. When the contribution PC 6 is large we see that dimuon SIDIS (purple) data are nearly orthogonal to DIS HERA (green) data. As the projection is rotated, we can also see that DIS HERA (green) practically lies on a plane in this 6D subspace. When the contribution of PC6 is near zero, dimonSIDIS (purple) occupies the same space as the DIS HERA data. A dynamic version can be viewed at \url{https://nspyrison.netlify.com/thesis/discluster_manualtour_pc6/}.

>>>>

\abstract{%
The set of dynamic linear projections of multivariate data collectively known as \textbf{tours} provides an important tool for extending the dimensionality of visuals. The R package \textbf{tourr} offers a variety of path generators and geometric displays for conducting tours. This paper discusses an extension package, \textbf{spinifex}, that adds support for the path generation of manual tours and extends the display of tours to use with the animation packages, \textbf{plotly} and \textbf{gganimate}. Manual tours are used to explore the sensitivity of structure as the contributions of a manipulation variable are changed.
A recent paper \{Wang et al. (2018)\} visualizes the sensitivity of the hadronic experiments to nucleon structure. Sensitivity was characterized in non-linear 3D embeddings of the first 10 principal components. This research applies manual tours to this data showing that manual tours resolve information about cluster structure and identify outliers, information that is orthogonal to the original static viewing plane.
}

@@

A tour is a multivariate data analysis technique in which a sequence of linear (orthogonal) projections are viewed as an animation while the orientation of the projection basis is rotated across time. Each frame of the sequence corresponds to a small change in the projection for a smooth transition that perseveres continuity. This is a crucial tool extending the dimensionality of visualization which is a large boon as datasets and models exist in increasing cardinality. The package \textbf{tourr} (Wickham et al. 2011) builds a platform for generating tour paths and applying to a wide number of display geometrics in \textbf{base} graphics.

While there are numerous methods that generate tour paths, this research focuses on the manual tour. The manual tour was described in Cook and Buja (1997) and allows a user to control the projection coefficients of a selected variable in a 2D projection. The manipulation of these coefficients allows the analyst to explore how sensitive the projections structure is to these changes. This makes manual tours particularly useful once a feature of interest has been identified, for example, with the use of a guided tour (Hurley and Buja 1990). Rather than exploring the large phase space of the data, a guided tour selects a very specific path via projection pursuit(Diaconis and Freedman 1984), the optimization of an index function on the projection via a hill climbing algorithm. This allows guided tours to rapidly identify interesting projection features given the relatively large parameter-space. After a projection of interest is identified an analyst can then use the ``finer brush'' of the manual tour, by controlling the contributions of individual variables.

Spinifex utilizes two new animation packages, \textbf{plotly} (Sievert 2018) and \textbf{gganimate} (Pedersen and Robinson 2019), to display tours, manual or other saved tours. From a given projection, the user can choose which variable to control, and the animation sequence is generated to remove the variable from the projection, and then extend its contribution to be the sole variable in one direction. This allows the viewer to assess the change in structure induced in the projection by the variable's contribution.

@@

\begin{itemize}
  \item $\textbf{X}$, the data, an $n \times p$ numeric matrix to be embedded in two dimensions.
  \item $\textbf{B} = (B_1,~ B_2)$, any orthonormal projection basis set, $p \times 2$ matrix, describing the projection from $p$ to two dimensions
  \item $\textbf{e}$, a zero column vector of length $p$ with the $k-$th element set to one, where $k$ is the number of the variable to manipulate.
  \item $\theta$, the angle of in-projection-plane rotation, for example, on the reference axes.
  \item $\phi$, the angle of out-of-projection-plane rotation, coming into the manipulation space.
\end{itemize}

The algorithm primarily operates on projection bases and utilizes the data only when making a display. The projection space can be viewed at any point in the process by pre-multiplying the data and plotting the first two variables.

@@

The flea data, originally from Lubischew (1962), made available in \textbf{tourr} is used to illustrate the algorithm. The data contains 74 observations across 6 variables, physical measurements of the flea beetles. Each observation belonging to one of three species.

The data is defined. A basis set (ideally orienting to an interesting feature) must be provided as an initial orientation. One way to identify a projection containing interesting features is to apply a guided tour (Cook, Swayne, and Buja 2007). In a guided tour the projection sequence is selected by optimizing an index function via hill-climbing on the projection space. In this case, the holes index is selected and applied to standardized flea data. The holes index is maximized when the projected observations are furthest from the center. Figure @ref\{fig:step0\} shows a locally optimized projection of the data. The left panel displays the reference axes of the projection basis, a visual indication of the magnitude and direction each variable contributed to the projection. The right panel shows the data as projected through the basis set described by the reference axes (left). Data points are colored and given shape according to the species (while the guided tour was unsupervised with this information).

(ref:step0-cap) Basis reference axes (left) and projected data (right) of standardized flea data. Data points color and shape are mapped to beetle species. Basis identified by a holes-index guided tour. The variables \texttt{aede2} and \texttt{tars1} contribute mostly orthogonal to the other variables.

@@

{\centering \includegraphics[width=0.6\linewidth]{spinifex_paper_files/figure-latex/step0-1}

@@

Call \texttt{view\_basis()} on a basis to produce a \textbf{ggplot2} graphic similar to @ref(fig:step0). Projection space is always available for display via the matrix multiplication \(\textbf{X}_{[n,~p]} ~*~ \textbf{B}_{[p,~d]} ~=~ \textbf{P}_{[n,~d]}\).

\begin{Schunk}
\begin{Sinput}
# devtools::install_github("nspyrison/spinifex") # Development version
install.package("spinifex")
library("spinifex")

f_data  <- tourr::rescale(flea[,2:6])
f_path  <- save_history(f_data, guided_tour(holes()))
f_basis <- matrix(f_path[,, max(dim(f_path)[3])], ncol=2)
f_cat   <- factor(flea$species)

view_basis(basis = f_basis, 
           data = f_data,
           labels = colnames(f_data))
\end{Sinput}
\end{Schunk}

@@

In figure @ref(fig:step0) the contributions of the variables \texttt{tars1} and \texttt{aede2} are mostly orthogonal to the contributions of the other four variables. These two variables explain the variation of the data distinguishing the purple group from the rest of the sample. Select \texttt{aede2} as the manip var, the variable to be manipulated, as it has a larger contribution to the projection. The question that will be explored is how important the variable \texttt{aede2} is to the separation of the clusters.

@@

Initialize a zero vector \textbf{e} of length \emph{p}. Set the fifth element to one, as \texttt{aede2} is the fifth variable in the data, giving the manip var a full contribution in this dimension. Use the Gram-Schmidt process to orthonormalize the zero vector onto the basis yielding the 3D manipulation space, \textbf{M}.

\begin{align*}
  \textbf{e} &\leftarrow Orthonormalize_{GS}(\textbf{e}) ~~~ w.r.t.~the~basis \\
  &\leftarrow \textbf{e} - \langle \textbf{e},\textbf{B}_1 \rangle \textbf{B}_1 - \langle \textbf{e}, \textbf{B}_2 \rangle \textbf{B}_2 \\
  \\
  \textbf{M}_{[p,~3]} &= (\textbf{B}_1,\textbf{B}_2,\textbf{e})
\end{align*}

Adding this new dimension to our projection space allows for the coefficients of the manip var to be changed via rotation about the origin. For example, the ability to lift a piece of paper, rather than being constrained to a 2D plane. Orthonormalizing rescales the new vector while the projection down to 2D remains the original basis. Place the plane horizontally, with the new dimension extending vertically, with axes projecting back onto the reference axes. Figure @ref(fig:step2) illustrates this 3D manipulation space with the manip var highlighted (height of the other variables are not depicted.)

(ref:step2-cap) Manipulation space for controlling the contribution of \texttt{aede2} of standardized flea data. Basis selected by a holes-index optimized guided tour lies on the horizontal projection plane, shown in blue. The manip var axis, in red, extends into the vertical manipulation space, allowing the coefficients of the manip var to be changed by rotation around the origin.

@@

{\centering \includegraphics[width=0.6\linewidth]{spinifex_paper_files/figure-latex/step2-1}

@@

The representation in figure @ref(fig:step2) can be produced by calling the function \texttt{view\_manip\_space()}.

\begin{Schunk}
\begin{Sinput}
view_manip_space(basis = f_basis, 
                 manip_var = 4, 
                 labels = colnames(f_data))
\end{Sinput}
\end{Schunk}

@@

Imagine holding the manip var, the red axis, one end fixed to the origin. As it is controlled the manipulation space rotates about the origin, the projection onto the horizontal projection plane correspondingly moves. This is what happens in a manual tour. Generating a sequence of values for the horizontal and vertical, angles produces a path for the rotation of the manipulation space. This defines the (orthonormally-constrained) rotation on the coefficients of the variables.

For a radial tour fix the (horizontal) angle within the projection plane, \(\theta\), and define a sequence for the (vertical) angle coming out of the projection plane, \(\phi\), bringing the initial \texttt{XY} contributions of the manip var to a maximum and then to zero before returning to the initial position. Dynamic capture of user manipulation is typically performed directly on the projection plane (without depiction of the manipulation space.)

@@

\begin{itemize}
  \item[$c_\theta$] be the cosine of $\theta$
  \item[$c_\phi$]   be the cosine of $\phi$
  \item[$s_\theta$] be the sine of   $\theta$
  \item[$s_\phi$]   be the sine of   $\phi$
\end{itemize}

then
\begin{align*}
  \textbf{RM}_{[p,~3,~i]}
  &= \textbf{M}_{[p,~3]} ~*~ \textbf{R}_{[3,~3,~i]} \\
  &= \begin{bmatrix}
    M_{1,~1} & M_{1,~2} & M_{1,~3} \\
    M_{2,~1} & M_{2,~2} & M_{2,~3} \\
    \vdots   & \vdots   \\
    M_{p,~1} & M_{p,~2} & M_{p,~3}
  \end{bmatrix}_{[p,~3]}
    ~*~
  \begin{bmatrix}
    c_\theta^2 c_\phi s_\theta^2 &
    -c_\theta s_\theta (1 - c_\phi) &
    -c_\theta s_\phi \\
    -c_\theta s_\theta (1 - c_\phi) &
    s_\theta^2 c_\phi + c_\theta^2 &
    -s_\theta s_\phi \\
    c_\theta s_\phi &
    s_\theta s_\phi &
    c_\phi
  \end{bmatrix}_{[3,~3,~i]}
\end{align*}

A note on application: \(\phi\) is the angle relative to the initial value of \(\phi\), we find the transformation \(\phi_i\) - \(\phi_1\) useful to think about \(\phi\) relative to the basis plane. Additionally, the value of \(\phi\) may be out of phase by a factor of pi. If the manip variable doesn't move as expected these are the first places to check.

Figure @ref(fig:step3) illustrates a sequence with 15 projected bases, showing the reference axes on top the corresponding projected data points on the below. As a result, we can see that changes in the manip var controlled the distance between the purple cluster and the remaining sample, \texttt{aede2} is crucial in distinguishing this species. Tours are typically viewed as an animation such a dynamic version of this tour can be viewed online at \url{https://nspyrison.netlify.com/thesis/flea_manualtour_mvar5/}. The page may take a moment to load.

(ref:step3-cap) Radial manual tour manipulating \texttt{aede2} of standardized flea data. The contributions increase from its initial contribution to a full contribution to the projection before decreasing to zero and then returning to its initial value. The change in the projected data shows that \texttt{aede2} is important for distinguishing the purple species. An animated version can be viewed at \url{https://nspyrison.netlify.com/thesis/flea_manualtour_mvar5/}.

@@

Animations can be produced using the function \texttt{play\_manual\_tour()}. This function defaults to an HTML5 widget produced from \textbf{plotly}. The \texttt{render\_type} argument can be changed to \texttt{render\_gganimate} for exporting to .gif or .mp4 files.

\begin{Schunk}
\begin{Sinput}
angle_speed <- .26

play_manual_tour(data = f_data,
                 basis = f_basis, 
                 manip_var = 5, 
                 angle = angle_speed,
                 col = f_cat,
                 pch = f_cat)
\end{Sinput}
\end{Schunk}

@@

In an appeal to performance, the above operations are performed on the bases without the use of the larger datasets. After the bases are brought into the projection-space, however, it is helpful to observe them with data in the same space. Pre-multiply the data by basis frame bringing the data into the projection space.

@@

The \textbf{tourr} package utilizes \textbf{base} graphics for the display of tours. \textbf{spinifex} allows tours to be used in rendered in \textbf{plotly} as an HTML5 object or \textbf{gganimate} as .gif or .mp4 objects. Both of which build off \textbf{ggplot2} objects in internal functions. Sharing of animations is not trivial especially in print and static file formats such as .pdf. Even with the use of computers and dynamic file formats capturing the correct resolution, aspect, and display is challenging and many formats quickly bloat file sizes. Keep in mind hosting options and exporting functions from \textbf{plotly}, \textbf{gganimate} and \textbf{tourr}.

@@

Storing each data point for every frame of the animation is redundant. Just as operations are performed on the bases, so too should tour paths be stored as bases and a single instance of the data. Consider a radial manual tour, we can store the salient features in 3 bases, where \(\phi\) is at its starting, minimum, and maximum values. The frames in between can be interpolated by supplying angular speed. With the use of the \texttt{tourr::save\_history()} function, the target bases can be saved. From there geodesic interpolation can be used to populate the intermittent frames. This type of interpolation should not be used on manual tours, which have already been initialized into a 3D manipulation space where direct linear interpolation is appropriate.

@@

In a recent paper, Wang et al. (2018), the authors aggregate and visualize the sensitivity of hadronic experiments to nucleon structure. The authors introduce a new tool, PDFSense, to aid in the visualization of Parton distribution functions (PDF). The parameter-space of these experiments lies in 56 dimensions, \(\delta \in \mathbb{R}^{56}\), and are visualized as 3D subspaces of the 10 first principal components in linear (PCA) and non-linear (t-SNE) embeddings.

Using the same data, another study, Cook, Laa, and Valencia (2018), applies grand tours (Asimov 1985) to the same subspaces. Grand tours are dynamic subspace projections of high dimensional where frames and selected at random and linked with geodesically interpolation of the intermediate frames. Grand tours are able to better resolve the distribution shape of clusters, intra-cluster detail, and better outlier detection than the use of PDFSense \& TFEP (TensorFlow embedded projections). Before applying manual tours let's discuss the structure of the data.

The data has a hierarchical structure with top-level clusters; \texttt{DIS}, \texttt{VBP}, and \texttt{jet}. Each cluster is a particular class of experiments, each with many experimental datasets which, in turn, have many observations. In the consideration of data density, we conduct manual tours on subsets of the \texttt{DIS} and \texttt{jet} clusters. This explores the sensitivity of the structure to each of the variables in turn and we present the subjectively best and worst variable to manipulate for identifying dimensionality of the clusters and describing the span of the clusters.

@@

The \texttt{jet} cluster resides in a smaller dimensionality than the full set of experiments with four principal components explaining 95\% of the variation in the cluster (Cook, Laa, and Valencia 2018). The data within this 4D embedding is subset down to \texttt{ATLAS7old} and \texttt{ATLAS7new} to focus in on two groups with a reasonable number of observations that occupy different parts of the subspace. Radial manual tours controlling contributions from PC4 and PC3 are shown in figure @ref(fig:JetClusterGood) and figure @ref(fig:JetClusterBad) respectively. These components are selected to contrast the difference of information conveyed by touring different variables. Links to dynamic HTML5 animations controlling each of the four variables are also provided.

When manipulating PC4, there is clear difference in the parameter space spanned by the experiment types \texttt{ATLAS7new} and \texttt{ATLAS7old}. Yet, when PC3 is manipulated there is no indication that the diffent expiments probe different parameter space. Performing a radial manual tour on PC4 is more insightful than for PC3.

Radial manual tours manipulating each of the principal components in the \texttt{jet} cluster can be viewed by following the links: \href{https://nspyrison.netlify.com/thesis/jetcluster_manualtour_pc1/}{PC1}, \href{https://nspyrison.netlify.com/thesis/jetcluster_manualtour_pc2/}{PC2}, \href{https://nspyrison.netlify.com/thesis/jetcluster_manualtour_pc3/}{PC3}, and \href{https://nspyrison.netlify.com/thesis/jetcluster_manualtour_pc4/}{PC4}.

(ref:JetClusterGood-cap) A radial manual tour of PC4 within the \texttt{jet} cluster. Colored by experiment type: \texttt{ATLAS7new} in green and \texttt{ATLAS7old} in orange. When PC4 fully/negligably contributes to the projection \texttt{ATLAS7new} (green) spans the same space as the orange points. During the intermediate frames the \texttt{ATLAS7new} is compressed in the direction radial to PC4. A dynamic version can be viewed at \url{https://nspyrison.netlify.com/thesis/jetcluster_manualtour_pc4/}.

@@

(ref:JetClusterBad-cap) A radial manual tour of PC4 within the \texttt{jet} cluster. Colored by experiment type: \texttt{ATLAS7new} in green and \texttt{ATLAS7old} in orange. Data from \texttt{ATLAS7new} (green) spans the mostly the same space as \texttt{ALTLAS7old} (orange) with no evident differnce in cluster structure across varying contributions of PC3. A dynamic version can be viewed at \url{https://nspyrison.netlify.com/thesis/jetcluster_manualtour_pc3/}.

@@

@@

A different space is used to explore the \texttt{DIS} cluster; specfically the first six pricipal components, which explain 48\% of the variation contained within the aggregated data(Cook, Laa, and Valencia 2018). Radial manual tours are performed on PC6 and PC2 in figure @ref(fig:DISclusterGood) and figure @ref(fig:DISclusterBad) respectively.

The selection of the manip variable is important as the manipulation spaces convey substantially different information. The manual tour of PC6 offers information about the dimensionality, shape, and orientations of the differnet experiment classes. whereas manipulating the contributions of PC2 only shows a subset of the dimensionality and shape information. Manipulating the contributions of PC6 turned out to be much more insightful than that of PC2. This result might seem counter-intuitive at first as PC2 should explain much more of the variation in the data. However, features and structure in the data regularly reside in finer detail that can be lost when looking only at static projections.

\texttt{DIS} cluster manual tours manipulating each of the principal components can be viewed from the links: \href{https://nspyrison.netlify.com/thesis/discluster_manualtour_pc1/}{PC1}, \href{https://nspyrison.netlify.com/thesis/discluster_manualtour_pc2/}{PC2}, \href{https://nspyrison.netlify.com/thesis/discluster_manualtour_pc3/}{PC3}, \href{https://nspyrison.netlify.com/thesis/discluster_manualtour_pc4/}{PC4}, \href{https://nspyrison.netlify.com/thesis/discluster_manualtour_pc5/}{PC5}, and \href{https://nspyrison.netlify.com/thesis/discluster_manualtour_pc6/}{PC6}.

(ref:DISclusterGood-cap) A radial manual tour manipulating the contribution of PC6 within the \texttt{DIS} cluster. Points are colored by experiment type: \texttt{DIS\ HERA1+2} in green, \texttt{dimuon\ SIDIS} in purple, and \texttt{charm\ SIDIS} in orange. The cluster \texttt{DIS\ HERA1+2} (green) is distributed in a cross shaped plane, \texttt{charm\ SIDIS} (orange) occupies the the center space of this cross, with the plane projecting into the field of view when the contribution of PC6 is max. Less evident is the linear \texttt{dimuon\ SIDIS} (purple) observations approaching the line of view for intermediate values of PC6. A dynamic version can be viewed at \url{https://nspyrison.netlify.com/thesis/discluster_manualtour_pc6/}.
